Gaia's constraints on the very massive (initially $\sim$$150\,M_\odot$) runaway star candidate VFTS682

### 2018-06-23 VFTS682_APJL.tex
selma's commits

% \title{\emph{Gaia} DR2 identifies VFTS 682 as the most massive % , dynamically
%                                 % ejected %% to fit in one line
% runaway star known to date}
%\title{Gaia's constraints on the  $\sim$$150\,M_\odot$  runaway star candidate VFTS682}
\title{Gaia's constraints on the extremely massive  ($\sim$$150\,M_\odot$) runaway star candidate VFTS682}

\author{M.~Renzo$^{*,1}$, S.~E.~de~Mink$^{1}$, D.~J.~Lennon$^{2}$,
  R.~P.~van~der~Marel$^{3,4}$, I.~Platais$^{4}$, 
  J.~Bestenlehner$^{5}$, C.~J.~Evans$^{6}$,
  V.~H\'enault-Brunet$^{7}$,  S.~Justham$^{1,8,9}$,  A.~de~Koter$^{1}$,
  N.~Langer$^{10}$, F. Najarro$^{11}$, H.~Sana$^{12}$, F.~R.~Schneider$^{13}$, J.~S.~Vink$^{14}$}
\affil{ $^{*}$ Corresponding author, \href{mailto:m.renzo@uva.nl}{m.renzo@uva.nl}.  A list of the affiliations can be found at the end of this paper.}

@@

       \hline
       \multicolumn{5}{l}{\emph{Selected stellar parameters VFTS 682}}
      \\
      \hline
     & present day mass  & $[M_\odot]$ & $137.8^{+27.5}_
                                           {-15.9}$ & (1)
                                                    \\
      & initial mass& $[M_\odot]$ & $150.0^{+28.7}_{-17.4}$ & (1)
      \\
      &age & [Myr] & $1.0\pm0.2$ & (1) \\
      &mass loss rate & [$M_\odot \ \mathrm{yr}^{-1}$] & $10^{-4.1\pm0.2}$ & (2)\\
      % &\multirow{2}{*}{surface helium mass fraction} &  & 0.45 & (2)\\
      % & &  & 0.49 & (3)\\ %% we don't use this: if we believe it's a
      % merger it's not relevant, if we don't we still don't use it
      \hline

    \end{tabular}
    \tablecomments
    { 
      (1)~\cite{schneider:18}
      (2)~\cite{bestenlehner:11}
      % (3)~\cite{rubio-diez:17}
    }
  \end{center}
  \label{tab:star_param}
\end{table}

\begin{table}
  \begin{center}
    \caption{Kinematics of VFTS682. }
    \begin{tabular}{llc|c|c}
      \hline
      \hline
      &Parameter & Units & Value & Ref.\\
      \hline
      \multicolumn{5}{l}{\emph{Coordinates}} \\
      \hline
         &$\mathrm{RA}_\mathrm{VFTS682}$&[degrees] & \phantom{-}84.7314 %$ \pm 0.036\,\mathrm{mas} $ 
                     & (1) \\        
               &$\mathrm{DEC}_\mathrm{VFTS682}$&[degrees] & -69.0741%$ \pm 0.048\,\mathrm{mas} $
                     & (1)  \\

                        &$\mathrm{RA}_\mathrm{R136}$&[degrees] & \phantom{-}84.6767
                     &  SIMBAD  \\        
               &$     \mathrm{DEC}_\mathrm{R136}$&[degrees] &  -69.1009
                     &  SIMBAD \\       
        &$      \delta\mathrm{RA}$  &[degrees] & \phantom{-0}0.0547                      
        &  (2, 5)
  \\        
               &$     \delta\mathrm{DEC}$  &[degrees] & \phantom{-0}0.0268 
                     &  (2, 5) \\

       \hline
           \multicolumn{5}{l}{\emph{Gaia proper motion for VFTS682
      and average of the region}} \\
      \hline
          &$\mu_\mathrm{RA}$&[$\masyr$] & $1.843\pm 0.070$
                     & (1) \\        
               &$\mu_\mathrm{DEC}$&[$\masyr$] & $0.786\pm 0.080$
                     &  (1) \\        
                 & $\rho\,(\mu_\mathrm{RA}, \mu_\mathrm{DEC})$ &[$\masyr$] & $0.026$
                        & (1)  \\         
     %  \multicolumn{5}{l}{\emph{Gaia's average proper motion in the region}} \\         
       &$\langle\mu_\mathrm{RA}\rangle_\mathrm{R136}$&[$\masyr$] & $1.74\pm0.01$
                        & (3) \\
      &$\langle\mu_\mathrm{DEC}\rangle_\mathrm{R136}$&[$\masyr$]
                & $0.70\pm0.02$ &  (3)\\
\hline

@@

      \multicolumn{5}{l}{\emph{Gaia relative proper motion for VFTS682 }} \\
      \hline
      &$\delta\mu_\mathrm{RA}$  &[$\masyr$] & $0.103\pm0.080$ & (1,5) \\
      &$\delta\mu_\mathrm{DEC}$  &[$\masyr$] & $0.086\pm0.100$ &  (1,5) \\
      &$\delta\mu_{Gaia}$  &[$\masyr$] & $0.13\pm0.09$ &  (1,5) \\
      &$v_\mathrm{2d}$  &[$\kms$] & $32\pm21$ & (1,5)\\  
      &$\theta$  &[degrees] &  $14_{-43}^{+44}$  & (1,5)\\  
      % &$v_\mathrm{2d, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
 %

 \hline     
      \multicolumn{5}{l}{\emph{HST relative proper motion for VFTS682}} \\
            \hline
      &$\delta\mu_\mathrm{RA, HST}$  &[$\masyr$] & $0.01\pm0.13$ & (4) \\
      &$\delta\mu_\mathrm{DEC, HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
       &$\delta\mu_\mathrm{HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
                  &$v_\mathrm{2d, HST}$  &[$\kms$] & $47\pm24$ & (4)\\  
                      &$\theta^{*}$  &[degrees] &   $-35_{-51}^{+24}$   & (1,5)\\  
                          % &$v_\mathrm{2d, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
 %

%\hline
     % \multicolumn{5}{l}{\emph{ Direction and magnitude w.r.t the vector pointing away from R136}} \\     
 %      &$\theta^{*}$  &[degrees] &  \SdM{xxx}$\pm$\SdM{xxx}  & (1,4)\\  
     %   &$v_\mathrm{2d, \parallel}$  &[$\kms$] & \SdM{xxx}$\pm$\SdM{xxx} & (1,4)\\  
 %          &$v_\mathrm{2d}$  &[$\kms$] & \SdM{xxx}$\pm$\SdM{xxx} & (1,4)\\

@@

The WNh5 star VFTS682, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) \mbox{-69$^\mathrm{o}$04'26.72''} J2000, was observed as part of the multi-epoch, spectroscopic VFTS campaign covering $\lambda$4000--7000 \citep[][]{evans:11}. 
%
%of spectral type WNh5 using multi-epoch spectra covering $\lambda$4000--7000 from the
%VFTS survey. % The spectra available covered $\lambda$4000--7000 and
% were taken at multiple epochs.
%
\citet{bestenlehner:11}  analyzed the spectra to infer the stellar parameters and measure an extinction of $A_V=4.45\pm0.12$, implying a luminosity of $\log_{10}(L/L_\odot) =  6.5\pm0.2$, making this one of the brightest stars in the region. The star is unlikely to have a close companion, because of the absence of clear radial velocity variations \citet{bestenlehner:11}.  Bayesian fits of the stellar parameters against evolutionary tracks \citep{brott:11, kohler:15} using the Bonsaii code \citep{schneider:17, schneider:18} provide estimates for the age, present mass and initial mass, indicating that this star among the most massive stars known, see \Tabref{tab:star_param} for an overview of the parameters.

Estimates of the radial velocity are complicated by the variable,
optically thick winds that is typical for emission line
stars. \citet{bestenlehner:11} estimate a mass loss rate of
$10^{-4.1\pm0.2}\,M_\odot \ \mathrm{yr}^{-1}$, not accounting for the
possible effect of clumping.  The V-band light curve of VFTS682  shows
variations at a $\sim$10\% level on a timescale of years, which is
unusual for Wolf-Rayet stars and more typical for Luminous Blue
Variable (LBV) stars \citep{udalski:08, bestenlehner:11}. It also
shows mid-infrared excess \citep{gruendl:09}.  We therefore caution for over-interpreting the estimates for the radial velocities, but we will summarize the estimates that have been published.  \citet{bestenlehner:11}  estimate a radial velocity (RV) of  $300\pm10\kms$ using the  N{\footnotesize V} $\lambda4944$ line, which is offset from the average radial velocity of the region, which is  $270\pm10\kms$. This is consistent with, but no proof of, a runaway nature.  \cite{bressert:12} note an offset compared between the RV of the star and the nebular lines from the gas filaments in its vicinity. This is consistent with, but also no proof of, the expectation if the star was not formed in situ.  Newer measurements based on new Xshooter data give different estimates for the RV ({\color{blue} Rubio-D{\' i}ez et al. in prep}). We will therefore refrain from using the RV measurements in this work.

%of $137.8^{+27.5}_ {-15.9}\Msun$, an apparent age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial mass of $150.0^{+28.7}_{-17.4}\Msun$.  Further parameters of are summarized in Table~\ref{X}.
VFTS682 is relatively isolated in \emph{Spitzer} images, except for the extended
source S10 interpreted as star formation triggered by the wind of
VFTS682 itself by \cite{walborn:13}. In X-rays
\todo{questions: does 682 show excess X-rays (wind collisions?)}

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}
  \includegraphics[width=0.35\textwidth]{figures/angle}
  \caption{\emph{Left panel}: distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136. VFTS682 is not an outlier, but
    its relative proper motion matches the expected value if it were indeed
    ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. The top axis shows the conversion to physical units
    assuming a distance of 50\,kpc. \emph{Right panel}:  distribution of
    angles between the relative proper motion direction and the radial
    direction from the center of the cluster to the star. The error bars from \emph{Gaia} for VFTS682 are large, but
    the best value is in agreement with the hypothesis of dynamical
    ejection. In both
    panels, the dark blue histograms contain 317 
    stars with error smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with errors smaller than $0.05\,\mathrm{mas \
      yr^{-1}}$.\todo{possibly: cartoon to illustrate angle definition, or add to
    fig.2 if not too messy, maybe also show HST measurements on this
    plot for 682? Redo right panel taking care of projection of the
    radial directions!It's misleading!}}
  \label{fig:dist}
\end{figure*}

@@

To determine the relative proper motion with respect to R136, we follow  \citet{lennon:18} to define the motion of the local frame of reference using the average proper motion of nearby stars with reliable astrometric data.  They select  bright ($G<17$) stars within 0.05 degrees of R136 and exclude sources with proper motion errors greater than $0.01\masyr$ in both coordinates (see their  Sect.~2.1 for further discussion).

With these we compute the relative proper motion
$\delta\mu_\mathrm{RA}$ and $\delta\mu_\mathrm{DEC}$.  We also derive
the angle $\theta$ between the direction of motion and the vector
connecting the center of R136 with the current position of VFTS 682,
i.e. such that $\theta = 0$ is the angle corresponding to perfectly
radial motion away from R136.    We further provide % the 2D velocity
% projected onto the radial vector pointing away from R136
% $v_\mathrm{2d, \parallel}$  and
the %total
projected 2d velocity
$v_\mathrm{2d}$.   All kinematic quantities are provided in
Tab.~\ref{tab:vfts682}. %%what does the projection on the proper
                        %%motion in the radial direction tell us?
                        %%Seems like a useless number to me...

@@

 \citet{platais:18} recently analyzed the HST data for candidate
 OB-type runaway stars. The brightest stars (V$<$14) are missing from
 the catalogue because of saturation effects, but the high extinction
 towards VFTS682 makes it faint enough to be included.
 This star was not included in their study because of its WNh5
 spectral type and because it would not pass the more conservative
 selection criteria they adopted.
 However, the measurements are useful here, since they provides an
 estimate of the proper motion of VFTS 682 that is completely independent from the \emph{Gaia} measurement.

We list their data in \Tabref{tab:vfts682}.

@@

@@

The left panel shows the distribution in two-dimensional proper motion relative
to R136, and shows that VFTS682 is not a peculiar star in terms of
its motion relative to R136. Subtracting the mean motion of the field,
we obtain $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
$\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$. Both indicate
a non-zero peculiar projected motion that would make it a slow runaway
star as expected by \cite{bestenlehner:11} if the star was ejected
from the cluster,  but in both cases the meaurements are also consistent with zero within 2$\sigma$. 
More importantly, both the proper motion
measurement from \emph{Gaia} and HST fall in the range expected to
reach its present day location % at $\sim$29\,pc from the cluster core
if the
star was ejected very early in its evolution. This range is
highlighted by the green vertical band, and its width is determined by
the uncertainty in the present-day age of the star (cf.~\Tabref{tab:star_param}).

The right panel shows the distributions in flight direction for these
stars: $\theta$ is the angle between the projected radial direction from the
center of R136 to each star and its proper motion relative to
R136 (see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce
close to radial ejections, i.e.~$\theta\simeq0$. On this panel, the green band highlights
the range of angles corresponding to an opening angle around the
radial direction of 45\,degrees. The direction of motion of VFTS682 is
poorly constrained because of the large errors on the proper motion
components both in HST and \emph{Gaia} data, but
it is consistent with dynamical ejection from R136. We note that
VFTS72 has a small radial velocity, while VFTS16 (and possibly VRTS682) has
large peculiar radial velocity and therefore accurate distances are
needed to contrain the flight direction in three dimensions.

Figure~\ref{fig:main} shows the projected motion of VFTS682 relative to R136 on the
sky. \todo{fix colors} The dashed circle shows the location of the cluster, and the red
(cyan) arrow shows the relative proper motion from \emph{Gaia} DR2
(HST). The lighter arrows show the uncertainty on the direction of
motion due to the large uncertainties. The length of each arrow is proportional to
the speed of the star, and dashed prolungations show the possible
region of origin assuming an age of $1.0\pm0.2$\,Myr
\citep[][]{schneider:18}. This figure illustrates that R136 is clearly
the most likely origin of the star, although the large error bars
prevent a robust identification.

Assuming VFTS682 indeed originates from R136, we can calculate its kinematic
age simply as:
\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{\delta\mu_{Gaia}} \simeq
  \frac{119.4\,\mathrm{arcsec}}{0.13\masyr} \simeq 0.9\pm\,0.6\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel = 119.4\,\mathrm{arcsec}$ is the angular distance from VFTS682 to
the core of the cluster \citep[corresponding to $\sim$29\,pc at LMC distance,][]{bestenlehner:11}.
As in the rest of this study, we neglect for
simplicity the error on the distance estimates, because it is negligible compared to other uncertainties.
The kinematic age $\tau_\mathrm{kin}$ is compatible with a very early
ejection from the cluster, given is apparent age.
% of $1.0\pm 0.2$\,Myr 
% \citep{schneider:18}.

@@

Based on our results, we tentatively claim that VFTS682 is the most massive
(slow) runaway known to date, with a peculiar two-dimensional
projected velocity with respect to R136 of
$\sim38\kms$ (averaging the \emph{Gaia} DR2 and HST
results) but compatible with zero within $2\sigma$. Due to the large error bars, this result will need
to be revisited with future astrometric data. % to be revisited as updated
% astrometric parameters from future \emph{Gaia} data releases become
% available.
If confirmed, it means that isolated star formation is
\emph{not} required to explain the isolation of VFTS682. Its proper motion suggests that it was ejected from the cluster R136
$0.9\pm0.6$\,Myr ago. Because of the exceptionally large mass
of this star, this raises the question of which stars must populate
the core of the cluster.

>>>>

% \title{\emph{Gaia} DR2 identifies VFTS 682 as the most massive % , dynamically
%                                 % ejected %% to fit in one line
% runaway star known to date}
%\title{Gaia's constraints on the  $\sim$$150\,M_\odot$  runaway star candidate VFTS682}
\title{Gaia's constraints on the very massive  (initially $\sim$$150\,M_\odot$) runaway star candidate VFTS682}

\author{M.~Renzo$^{*,1}$, S.~E.~de~Mink$^{1}$, D.~J.~Lennon$^{2}$, \SdM{I.~Platais$^{3}$,}
  R.~P.~van~der~Marel$^{3,4}$,  \SdM{E.~L.~Laplace$^{1}$,}
  J.~Bestenlehner$^{5}$, C.~J.~Evans$^{6}$,
  V.~H\'enault-Brunet$^{7}$,  S.~Justham$^{1,8,9}$,  A.~de~Koter$^{1}$,
  N.~Langer$^{10}$, F. Najarro$^{11}$, H.~Sana$^{12}$, F.~R.~Schneider$^{13}$, J.~S.~Vink$^{14}$}
\affil{ $^{*}$ Corresponding author: \href{mailto:M.Renzo@uva.nl}{M.Renzo@UvA.nl}.  A list of the affiliations can be found at the end of this paper.}

@@

       \hline
       \multicolumn{5}{l}{\emph{Selected stellar parameters VFTS 682}}
      \\
      \hline
     & present day mass  & $[M_\odot]$ & $137.8^{+27.5}_
                                           {-15.9}$ & (1)
                                                    \\
      & initial mass& $[M_\odot]$ & $150.0^{+28.7}_{-17.4}$ & (1)
      \\
      &age & [Myr] & $1.0\pm0.2$ & (1) \\
      &mass loss rate & [$M_\odot \ \mathrm{yr}^{-1}$] & $10^{-4.1\pm0.2}$ & (2)\\
      &\multirow{2}{*}{surface helium mass fraction} &  & 0.45 & (2)\\
      & &  & 0.49 & (3)\\
      \hline

    \end{tabular}
    \tablecomments
    { 
      (1)~\cite{schneider:18}
      (2)~\cite{bestenlehner:11}
      (3)~\cite{rubio-diez:17}
    }
  \end{center}
  \label{tab:star}
\end{table}

\begin{table}
  \begin{center}
    \caption{Kinematics of VFTS682. }
    \begin{tabular}{llc|c|c}
      \hline
      \hline
      &Parameter & Units & Value & Ref.\\
      \hline
      \multicolumn{5}{l}{\emph{Absolute and relative position}} \\
      \hline
         &$\mathrm{RA}_\mathrm{VFTS682}$&[degrees] & \phantom{-}84.7314 %$ \pm 0.036\,\mathrm{mas} $ 
                     & (1) \\        
               &$\mathrm{DEC}_\mathrm{VFTS682}$&[degrees] & -69.0741%$ \pm 0.048\,\mathrm{mas} $
                     & (1)  \\

                        &$\mathrm{RA}_\mathrm{R136}$&[degrees] & \phantom{-}84.6767
                     &  SIMBAD  \\        
               &$     \mathrm{DEC}_\mathrm{R136}$&[degrees] &  -69.1009
                     &  SIMBAD \\       
        &$      \delta\mathrm{RA}$  &[degrees] & \phantom{-0}0.0547                      
        &  (2, 5)
  \\        
               &$     \delta\mathrm{DEC}$  &[degrees] & \phantom{-0}0.0268 
                     &  (2, 5) \\  
                       &$  \mathrm{d_\mathrm{2d}}$  &[pc] & 29
                     &  (Besthenlehner) \\

                     \hline
           \multicolumn{5}{l}{\emph{Gaia absolute proper motion for VFTS682
      and the region}} \\
      \hline
          &$\mu_\mathrm{RA}$&[$\masyr$] & $1.843\pm 0.070$
                     & (1) \\        
               &$\mu_\mathrm{DEC}$&[$\masyr$] & $0.786\pm 0.080$
                     &  (1) \\        
                 & $\rho\,(\mu_\mathrm{RA}, \mu_\mathrm{DEC})$ &[$\masyr$] & $0.026$
                        & (1)  \\         
     %  \multicolumn{5}{l}{\emph{Gaia's average proper motion in the region}} \\         
       &$\langle\mu_\mathrm{RA}\rangle_\mathrm{R136}$&[$\masyr$] & $1.74\pm0.01$
                        & (3) \\
      &$\langle\mu_\mathrm{DEC}\rangle_\mathrm{R136}$&[$\masyr$]
                & $0.70\pm0.02$ &  (3)\\
\hline

@@

      \multicolumn{5}{l}{\emph{Gaia relative proper motion for VFTS682 }} \\
      \hline
      &$\delta\mu_\mathrm{RA}$  &[$\masyr$] & $0.103\pm0.080$ & (1,5) \\
      &$\delta\mu_\mathrm{DEC}$  &[$\masyr$] & $0.086\pm0.100$ &  (1,5) \\
      &$\delta\mu_{Gaia}$  &[$\masyr$] & $0.13\pm0.09$ &  (1,5) \\
      &$v_\mathrm{2d}$  &[$\kms$] & $32\pm21$ & (1,5)\\  
      &$\theta$  &[degrees] &  $14.2_{-43}^{+44}$  & (1,5)\\  
%      &$v_\mathrm{2d, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
 %

 \hline     
      \multicolumn{5}{l}{\emph{HST relative proper motion for VFTS682}} \\
            \hline
      &$\delta\mu_\mathrm{RA, HST}$  &[$\masyr$] & $0.01\pm0.13$ & (4) \\
      &$\delta\mu_\mathrm{DEC, HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
       &$\delta\mu_\mathrm{HST}$  &[$\masyr$] & $0.20\pm0.10$ &  (4) \\
                  &$v_\mathrm{2d, HST}$  &[$\kms$] & $47\pm24$ & (4)\\  
                      &$\theta^{*}$  &[degrees] &   $-35_{-51}^{+24}$   & (1,5)\\  
                  %        &$v_\mathrm{2d, \parallel}$  &[$\kms$] & $\SdM{xxx}\pm\SdM{xxx}$ & (1,5)\\  
    \hline
       \multicolumn{5}{l}{\emph{Expected proper motion if ejected from R136 at zero age}} \\
      \hline
      &$v_\mathrm{2d}$  &[$\kms$] & \SdM{$xx\pm xx$} & (Bestenlehner) \\  
      &$\theta$  &[degrees] &  $0$  & \\ 
       \hline

@@

The WNh5 star VFTS682, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) \mbox{-69$^\mathrm{o}$04'26.72''} J2000, was observed as part of the multi-epoch, spectroscopic VFTS campaign covering $\lambda$4000--7000 \citep[][]{evans:11}. 
%
%of spectral type WNh5 using multi-epoch spectra covering $\lambda$4000--7000 from the
%VFTS survey. % The spectra available covered $\lambda$4000--7000 and
% were taken at multiple epochs.
%
\citet{bestenlehner:11}  analyzed the spectra to infer the stellar parameters and measure an extinction of $A_V=4.45\pm0.12$, implying a luminosity of $\log_{10}(L/L_\odot) =  6.5\pm0.2$, making this one of the brightest stars in the region. The star is unlikely to have a close companion, because of the absence of clear radial velocity variations \citet{bestenlehner:11}.  Bayesian fits of the stellar parameters against evolutionary tracks \citep{brott:11, kohler:15} using the Bonsaii code \citep{schneider:17, schneider:18} provide estimates for the age, present mass and initial mass, indicating that this star among the most massive stars known, see \Tabref{tab:star} for an overview of the parameters.

Estimates of the radial velocity are complicated by the variable, optically thick winds that is typical for emission line stars. \citet{bestenlehner:11} estimate a mass loss rate of $10^{-4.1\pm0.2}\,M_\odot \ \mathrm{yr}^{-1}$, not accounting for the possible effect of clumping.  The V-band light curve of VFTS682  shows variations at a $\sim$10\% level on a timescale of years, which is unusual for Wolf-Rayet stars and more typical for Luminous Blue Variable (LBV) stars \citep{udalski:08, bestenlehner:11}. It also shows mid-infrared excess \citep{gruendl:09}.  We therefore caution for over-interpreting the estimates for the radial velocities, but we will summarize the estimates that have been published.  \citet{bestenlehner:11}  estimate a radial velocity (RV) of  $300\pm10\kms$ using the  N{\footnotesize V} $\lambda4944$ line, which is offset from the average radial velocity of the region, which is  $270\pm10\kms$. This is consistent with, but no proof of, a runaway nature.  \cite{bressert:12} note an offset compared between the RV of the star and the nebular lines from the gas filaments in its vicinity. This is consistent with, but also no proof of, the expectation if the star was not formed in situ.

Newer measurements based on a fit to new Xshooter data covering the full spectrum from 3200 Angstr{\"o}m to 2.45 microns lead to an estimate of $260 \pm 15 \kms$ ({\color{blue} Rubio-D{\' i}ez et al. in prep}). This estimate, which we consider to be the most reliable fo those quoted here, indicates that the relative motion along the line-of-sight is small or negligible. This in consistent with both hypotheses.

%of $137.8^{+27.5}_ {-15.9}\Msun$, an apparent age of $1.0^{+0.2}_{-0.2}$\,Myr and an inferred initial mass of $150.0^{+28.7}_{-17.4}\Msun$.  Further parameters of are summarized in Table~\ref{X}.

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}
  \includegraphics[width=0.35\textwidth]{figures/angle}
  \caption{\emph{Left panel}: distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136. VFTS682 is not an outlier, but
    its relative proper motion matches the expected value if it were indeed
    ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. The top axis shows the conversion to physical units
    assuming a distance of 50\,kpc. \emph{Right panel}:  distribution of
    angles between the relative proper motion direction and the radial
    direction from the center of the cluster to the star. The error bars from \emph{Gaia} for VFTS682 are large, but
    the best value is in agreement with the hypothesis of dynamical
    ejection. In both
    panels, the dark blue histograms contain 317 
    stars with error smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with errors smaller than $0.05\,\mathrm{mas \
      yr^{-1}}$.\todo{possibly: cartoon to illustrate angle definition, or add to
    fig.2 if not too messy}}
  \label{fig:dist}
\end{figure*}

@@

Gaia provides absolute proper motions.  To determine the relative proper motion with respect to R136, we follow  \citet{lennon:18} to define the motion of the local frame of reference using the average proper motion of nearby stars with reliable astrometric data.  They select  bright ($G<17$) stars within 0.05 degrees of R136 and exclude sources with proper motion errors greater than $0.01\masyr$ in both coordinates (see their  Sect.~2.1 for further discussion).

With these we compute the relative proper motion
$\delta\mu_\mathrm{RA}$ and $\delta\mu_\mathrm{DEC}$.   We also compute the total projected 2d velocity
$v_\mathrm{2d}$ and
the angle $\theta$ between the direction of motion and the vector
connecting the center of R136 with the current position of VFTS 682.  This means that $\theta = 0$ is the angle that is expected for a star that is moving away from R136 with a perfect radial motion.   All kinematic quantities are provided in
Tab.~\ref{tab:vfts682}.

@@

 \citet{platais:15, platais:18} analyzed the HST data to determine the relative proper motion and identify candidate runaway stars. The brightest stars (V$<$14) are saturated in the data set and have been excluded from the analysis.  The very massive star VFTS682 should be one of the brightest stars in the region, the high extinction makes it faint enough such that saturation effects are not an issue.   VFTS682 is relatively slow and is therefore not discussed in their paper.  However, their measurements are useful here, since they provides an  estimate of the proper motion of VFTS 682 that is independent from the \emph{Gaia} measurement.

The HST provides proper motions that are relative to the bulk motion of the majority of the stars in the field of view. The full 30 Dor field is covered by different pointings and there is some systematic distortion.  However, even for stars far from 30 Dor the effect is small, no more than $0.05\masyr$ across the whole 30 Dor field. The effect is much smaller for stars close to the center of field, such as VFTS 682 \citet{platais:18}.  We can therefore use the relative proper motion as a good estimate for the proper motion relative to R136.

 We list the relative proper motion in \Tabref{tab:vfts682}.

@@

At first glance the magnitude of the relative proper motions derived by Gaia,  $32\pm 21\kms$, and by HST, $47 \pm 24\kms$, appear consistent.  In principle they are consistent with the velocity

@@

The left panel shows the distribution in two-dimensional proper motion relative
to R136, and shows that VFTS682 is not a peculiar star in terms of
its motion relative to R136. Subtracting the mean motion of the field,
we obtain $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
$\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$. Both indicate
a non-zero peculiar projected motion that would make it a slow runaway
star as expected by \cite{bestenlehner:11} if the star was ejected
from the cluster,  but in both cases the meaurements are also consistent with zero within 2$\sigma$. 
More importantly, both the proper motion
measurement from \emph{Gaia} and HST fall in the range expected to
reach its present day location % at $\sim$29\,pc from the cluster core
if the
star was ejected very early in its evolution. This range is
highlighted by the green vertical band, and its width is determined by
the uncertainty in the present-day age of the star (cf.~\Tabref{tab:star}).

The right panel shows the distributions in flight direction for these
stars: $\theta$ is the angle between the radial direction from the
center of R136 to each star and its proper motion relative to
R136 (see also \Figref{fig:main}). Dynamical ejections from the
cluster should %are expected to
produce
close to radial ejections, i.e.~$\theta\simeq0$. On this panel, the green band highlights
the range of angles corresponding to an opening angle around the
radial direction of 45\,degrees. The direction of motion of VFTS682 is
poorly constrained because of the large errors on the proper motion
components both in HST and \emph{Gaia} data, but
it is consistent with dynamical ejection from R136.

Figure~\ref{fig:main} shows the projected motion of VFTS682 relative to R136 on the
sky. \todo{fix colors} The dashed circle shows the location of the cluster, and the red
(cyan) arrow shows the relative proper motion from \emph{Gaia} DR2
(HST). The lighter arrows show the corresponding opening angle within
the large uncertainties. The length of each arrow is proportional to
the speed of the star, and dashed prolungations show the possible
region of origin assuming an age of $1.0\pm0.2$\,Myr
\citep[][]{schneider:18}. This figure illustrates that R136 is clearly
the most likely origin of the star, although the large error bars
prevent a robust identification.

Assuming VFTS682 originates from R136, we can calculate its kinematic
age simply as:
\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{\delta\mu_{Gaia}} \simeq
  \frac{119.4\,\mathrm{arcsec}}{0.13\masyr} \simeq 0.9\pm\,0.6\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel = 119.4\,\mathrm{arcsec}$ is the angular distance from VFTS682 to
the core of the cluster \citep[corresponding to $\sim$29\,pc at LMC distance,][]{bestenlehner:11}.
As in the rest of this study, we neglect for
simplicity the error on the distance estimates, because it is negligible compared to other uncertainties.
The kinematic age $\tau_\mathrm{kin}$ is compatible with a very early
ejection from the cluster, given is apparent age.
% of $1.0\pm 0.2$\,Myr 
% \citep{schneider:18}.

@@

Based on our results, we tentatively claim that VFTS682 is the most massive
(slow) runaway known to date, with a peculiar three-dimensional speed of
$44\pm21\,\kms$ \todo{make 2d? use LOS velocity from Marimar?}. Due to the large error bars, this result will need
to be revisited with future astrometric data. % to be revisited as updated
% astrometric parameters from future \emph{Gaia} data releases become
% available.
If confirmed, it means that isolated star formation is
\emph{not} required to explain the isolation of VFTS682. Its proper motion suggests that it was ejected from the cluster R136
$0.9\pm0.6$\,Myr ago. Because of the exceptionally large mass
of this star, this raises the question of which stars must populate
the core of the cluster.
